Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-18 icfp.tex
RM edits metatheory section
And, especially, completely rewrites the reverse binder swap section.

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
A main goal of the Linear Core type system is 
to serve as a typed intermediate representation for optimising
compilers of lazy languages with linearity. In light of this,
we show that multiple optimising transformations are type
preserving in Linear Core (i.e.,~preserve linearity semantically).

@@

We now discuss an issue we have found with the so-called reverse binder swap
transformation.
%
%
Reverse binder swap is dual to the binder swap transformation,
substituting occurrences of the case
binder $z$ by the scrutinee in the special case when the scrutinee is a
variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By
using the scrutinee $x$ instead of the case binder, we might be able to float
out expressions from the alternative using the case binder. For example, we
might float an expensive computation involving $z$ out of the case alternative,
where $z$ is now out of scope:

@@

Despite GHC applying the reverse binder swap transformation 
during its Core-to-Core optimisation passes, this optimisation actually \emph{breaks} linearity
when considered as a transformation on Linear Core programs. In practice, the
optimisation preserves linearity when applied as part of the GHC
pipeline due to the occurrence analyser being naive with
regard to semantic linearity. Initially, it might seem as though an expression
in which a variable $x$ occurs both in the case scrutinee and in the
alternatives is linear (e.g.,~$\ccase{x}{\_ \to x}$):
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
  If $x$ refers to an unevaluated expression, then scrutinizing it results
in the expression bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
%
On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables).
%\end{itemize}
%
% This idea would be encoded by a possible rule like $$\TypeCaseVar$$
%

Even though, on its own, it makes intuitive sense that this example indeed uses
$x$ linearly, when considered as part of a complete type system, allowing this
expression to be linear makes the system \emph{unsound}.
To see why this is the case, recall that $\beta$-reduction reduces an application of a linear function
using call-by-name -- if we know the argument is used exactly once, a binding
to share the result of computing the argument is unnecessary, so we instead
substitute the argument expression for the linearly-bound variable in the
$\lambda$-body directly.

Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
where $y$ is a free linear variable. Assuming $\lambda x.~\ccase{x}{\_ \to x}$
is a linear function, $\beta$-reduction transforms the
application into $\ccase{use~y}{\_ \to
  use~y}$. Now, $y$ is a linear variable \emph{consumed} in the scrutinee of the case
that also occurs in the case alternative --
linearity is violated by using the linear resource $y$ twice.
%
% Essentially, Linear Core would be unsound, and even duplicate resources, if the
% above kind of expressions, where linear variable scrutinees occur in the
% alternatives body, were well-typed, because of its interaction with the
% call-by-name reduction of linear functions.
In this sense, the reverse binder
swap is an optimisation that creates ill-typed expressions from well-typed
ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
our system.

Reverse binder swap is not a problem in the GHC simplifier because of the
weaker notion of linearity understood by occurrence analysis. Occurrence
analysis is a static analysis pass which can be used to determine whether a
lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
$\beta$-reduction is done with call-by-need on such an expression. If the above
example were reduced with call-by-need
% \[
% \begin{array}{l}
% (\lambda x.~\ccase{x}{\_ \to x})~(use~y)
% \Longrightarrow_\textrm{CBNeed}
% \llet{x = use~y}{\ccase{y}{\_ \to y}}
% \end{array}
% \]
then the computation using $y$ would be let-bound, and $y$ used as a scrutinee
variable, which is indeed semantically linear in $x$.

@@

    \item Programs which \emph{actually violate linearity}, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to be able to reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource, and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199,10.1007/978-3-642-24276-2_2}, and
        others too, could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system of insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.

>>>>

\subsection{Optimisations preserve linearity\label{sec:optimisations-preserve-types-meta}}
%\todo[inline]{Check Commuting case-let in Fig.~\ref{fig:opttrans}!}
A main goal of the Linear Core type system is 
to serve as a typed intermediate representation for optimising
compilers of lazy languages with linearity. In light of this,
we show that multiple optimising transformations are type
preserving in Linear Core. %(i.e.,~preserve linearity semantically).

@@

Reverse binder swap is dual to the binder swap transformation, substituting
occurrences of the case binder $z$ by the scrutinee in the special case when
the scrutinee is a variable $x$.
%
% his is dual to what the binder swap transformation
% does in hope of eliminating multiple uses of $x$ (allowing for inlining). 
By using the scrutinee $x$ instead of the case binder, we might be able to
float out expressions from the alternative using the case binder. For example,
we might float an expensive computation involving $z$ out of the case
alternative, where $z$ is now out of scope:

@@

Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, \emph{this optimisation violates linearity} when considered
as a transformation \emph{on Linear Core programs} because the result (second program
above) is ill-typed -- the resource $x$ of the scrutinee not in WHNF is used
again in an alternative, despite it no longer being available according to our system.

An expression in which a variable $x$ occurs both in the case scrutinee and in
the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
If $x$ refers to an unevaluated expression, then scrutinizing will force the expression
bound by $x$ to be evaluated to WHNF. In a subsequent use of
$x$ in the alternatives, $x$ refers to the evaluated scrutinee in WHNF, which
must be consumed. Since $x$ is just another name (like the case binder) for the
scrutinee in WHNF, we may use it instead of the case binder or pattern
variables.
%
On the other hand, if $x$ already refers to an expression evaluated to WHNF, then
scrutinizing it in a case expression is a no-op, thus we may use it again (in
mutual exclusion with the case binder and pattern variables).

However, allowing this expression to be linear makes $\lambda^\pi_\Delta$ unsound!
Consider the function application $(\lambda x.~\ccase{x}{\_ \to x})~(use~y)$
where $y$ is a free linear variable, and assume $\lambda x.~\ccase{x}{\_ \to x}$
to be a linear function. According to $\lambda^\pi_\Delta$'s call-by-name
$\beta$-reduction, we have:
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBName}
\ccase{use~y}{\_ \to use~y}
\end{array}
\]
After the reduction, $y$ is a linear variable \emph{consumed} in the scrutinee
of the case that also occurs in the case alternative -- the resource
$y$ is being effectively duplicated.
%
If the application were instead reduced with call-by-need
\[
\begin{array}{l}
(\lambda x.~\ccase{x}{\_ \to x})~(use~y)
\Longrightarrow_\textrm{CBNeed}
\llet{x = use~y}{\ccase{x}{\_ \to x}}
\end{array}
\]
then the computation using $y$ would be let-bound, and $y$ would be used as a
scrutinee variable, which preserves semantic linearity.

For an optimising compiler such as GHC, the reverse binder swap transformation
violating linearity in Linear Core is relevant because optimising
compilers do not necessarily allocate a binding/use call-by-need for
every function application -- our result shows that failing to do so
when calling a function such as the one above is \emph{unsound}.
%
In an optimising compiler, a binding may be deemed unnecessary if the function
argument is inferred to be used exactly once. Linearity can play a
(non-heuristic) role in this inference, for a linear function uses its
argument exactly once by definition.
%
% Since the compiler may leverage this information (we
% briefly discuss \emph{linearity-influenced optimisations} in
% Section~\ref{sec:discussion}).
%
However, it is incompatible to both type expressions like $(\lambda
x.~\ccase{x}{\_ \to x})$ as linear functions and consider
call-by-name $\beta$-reduction of linear functions unconditionally
safe (without risk of duplicating resources nor work).
%
Furthermore, considering such an use of $x$ to be linear also makes inlining of
linear let bindings no longer unconditionally safe.

% $\beta$-reduction reduces an application of a linear function using
% call-by-name -- If we know the argument is used exactly once, a binding to
% share the result of computing the argument is unnecessary, so we instead
% substitute the argument expression for the linearly-bound variable in the
% $\lambda$-body directly.
%%
%% Essentially, Linear Core would be unsound, and even duplicate resources, if the
%% above kind of expressions, where linear variable scrutinees occur in the
%% alternatives body, were well-typed, because of its interaction with the
%% call-by-name reduction of linear functions.
%In this sense, the reverse binder
%swap is an optimisation that creates ill-typed expressions from well-typed
%ones, so it is deemed an \emph{invalid} optimisation that does not preserve types in
%our system.

%% Reverse binder swap is not a problem in the GHC simplifier because of the
%% weaker notion of linearity understood by occurrence analysis. Occurrence
%% analysis is a static analysis pass which can be used to determine whether a
%% lambda application can be $\beta$-reduced call-by-name, and $\ccase{x}{\_ \to
%% x}$ is \emph{not} seen as using $x$ linearly by the analysis. Thus,
%% $\beta$-reduction is done with call-by-need on such an expression.
%
% In practice, the optimisation preserves linearity when applied as part of the
% GHC pipeline due to the occurrence analyser being naive with regard to semantic
% linearity.

@@

    \item Programs which \emph{actually violate linearity}, in that the term
        violates the linear type it is annotated with. These are the programs
        we ultimately want to be able to reject by having a linearly typed Core
        validate intermediate programs. We found both a program which outright
        discarded a linear resource, and that rewrite rules implementing stream
        fusion~\cite{10.1145/165180.165214,10.1145/1291151.1291199}, and
        others too, could produce linearly-invalid programs (in one such case,
        the type of \lstinline{build} was not general enough to accommodate linearity).
\end{itemize}
%
Overall, the programs rejected by the Linear Core plugin,
%, in \texttt{linear-base},
besides validating that our implementation is faithful to the
$\lambda^\pi_\Delta$ system of insofar as programs that should not be accepted
are deemed ill-typed, provide further insight into the remaining
details required to fully typecheck linearity in a mature optimising compiler.
